---
title: "Random Variables for Parasite Infection Dynamics"
subtitle: "Computing the MoI, AoI, and AoY"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Variables for Parasite Infection Dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(ramp.falciparum)
library(ramp.func)
suppressWarnings(library(viridis))
suppressMessages(library(viridisLite))
``` 

We have derived mathematical formulas that describe the dynamics of malaria infections as random variables in cohorts of humans as they age (Henry JM, *et al.*, 2024)^[Henry JM, Carter AR, Wu SL, Smith DL (in preparation). A Probabilistic Synthesis of Malaria Epidemiology: Exposure, Infection, Parasite Densities, and Detection.] This R package is the computational companion. 

In the following, we review the mathematical formulas and the functions in `ramp.falciparum` that compute the multiplicity of infection (MoI), the age of infection (AoI), and the age of the youngest infection (AoY). 


## Formulas and Computation

1. Let $z_d(\alpha,a)$ denote the density for infections of age $\alpha$ in a host cohort of age $a$ born on day $d$. 

    + The dynamics of $z(\alpha,a)$ are described by the following:
$$
\frac{\partial z}{\partial a} + \frac{\partial z}{\partial \alpha} = - r z
$$
with the boundary condition:
$$
\begin{equation}
    z_d(0,a) = h_d(a).
\end{equation}
$$
Its solutions are given by: 
$$
z_d(\alpha,a) = h_d(a-\alpha) e^{-r \alpha}
$$

    + The function `zda` computes $z_d(\alpha, a).$ 

2. The mean MoI is given by the formula: 
$$
m_d(a) = \int_0^a z_d(\alpha, a) d \alpha
$$
    + The distribution of the MoI is Poisson (Nåsell I, 1985)^[Nåsell I (1985). Hybrid Models of Tropical Infections, 1st edition. Springer-Verlag. https://doi.org/10.1007/978-3-662-01609-1] with mean $m_d(a).$ 
    
    +  The function `meanMoI` computes $m_d(a)$ using `zda`

3. The true prevalence is the 
$$
p_d(a) = 1 - e^{-m_d(a)} 
$$
The function `truePR` computes the true prevalence using `meanMoI` 

4. The density function for the age of infection (AoI) is
$$
A_d(a) \sim f_A(\alpha, a, d) = \frac{z_d(\alpha,a)}{m_d(a)}
$$
and its moments are
$$
x_n(a, d | h) = \int_0^a  \alpha^n f_A(\alpha, a, d) d \alpha
$$

5. The age of the youngest infection (AoY) is defined as: 
$$
Y_d(a) \sim f_Y(\alpha, a, d) = \min_{\zeta \sim M_d(a)} \left\{ \alpha_i \right\}_{i=1}^\zeta,  \alpha_i \sim A_d(a) 
$$

    + The density function can be expressed in terms of the density and distribution functions of the AoI and MoI. 
$$
f_Y(\alpha; a, d) = 
f_A(\alpha, a,d) e^{-m_d(a) F_A(\alpha, a,d)} \frac{m_d(a)}{p_d(a)}.
$$ 
    + The distribution function for the AoY is:
$$
F_Y(a) \sim
\frac{1-e^{-m_d(a)F_A(\alpha, a,d)}}{1-e^{-m_d(a)}} = 
\frac{1-e^{-m_d(a)F_A(\alpha, a,d)}}{p_d(a)} 
\label{FY}
$$

    + Its moments are:
$$
y_n(a, d | h) = \int_0^a  \alpha^n f_Y(\alpha | a, d, h) d \alpha 
$$
5. We also developed functions to compute the age of the youngest of $N$ infections, called YoN 
$$
N_d(a) \sim
\min_{N} \left\{ \alpha_i \right\}_{i=1}^N \mbox { where }  \alpha_i \sim A_d(a) 
$$
    + The distribution function for YoN, $N_d(a)$, is 
$$F_N(\alpha, a, t) \sim 1- (1-F_A(\alpha, a, d))^N$$
The following is a summary table of functions to compute the MoI, AoI, AoY, and all their moments.
    
    + The density function for YoN is found by differentiating:
    
$$f_N(\alpha, a, t) \sim N (1-F_A(\alpha, a, d))^{N-1}\frac{f_A(\alpha, a, d)}{m_d(a)}$$


## Quick Reference 

The following is a summary table of functions to compute the MoI, AoI, AoY, YoN, and all their moments.

<center>
| | MoI | AoI | AoY | YoN |
| ------ | :---: | :---: | :---: | :---: |
` `  | $\zeta$ |  $\alpha$ | $\alpha$ | $\alpha$ | 
` ` | $\zeta \geq 0$ |  $0 \leq \alpha \leq a$ | $0 \leq \alpha \leq a$ | $0 \leq \alpha \leq a$ | 
Random Variable | $M_d(\zeta, a, h)$ |  $A_d(\alpha, a, h)$ | $Y_d(\alpha, a , h)$ | $N_d(\alpha, a, h)$| 
Density Function | $f_M(\zeta, a, h)$ | $f_A(\zeta, a, h)$  | $f_Y(\zeta, a, h)$ |  $f_N(\zeta, a, h)$ | 
` ` | dpois | dAoI | dAoY | dYoN |  
Distribution Function | $F_M(\zeta, a, h)$ | $F_A(\zeta, a, h)$  | $F_Y(\zeta, a, h)$ | $F_N(\zeta, a, h)$ | 
` ` | ppois | pAoI | pAoY | pYoN | 
Random Numbers  | $\hat M_d(\zeta, a, h)$| $\hat A_d(\alpha, a, h)$ | $\hat Y_d(\alpha, a , h)$ | $\hat N_d(\alpha, a , h)$ | 
` ` | rpois | rAoI | rAoY | rYoN | 
Moments | $m_d(a, h)$ |  $x_n(a, d, h)$ | $y_n(a, d, h)$ |
` ` | meanMoI | momentAoI | momentAoY | 
</center>


## Demonstration


### Force of Infection (FoI)

```{r}
library(ramp.func)
# devtools::load_all()
# devtools::load_all("~/git/ramp.func")
```


```{r, fig.width=7, fig.height=4}
clrs = viridisLite::turbo(7)
set.seed(234)
Sa = makepar_F_type2()
Sp = makepar_F_sin()

F_t <- make_ts_function(scale = 1, season_par=Sp)
FoI_a <- make_F_a(avg = 3/365, age_par=Sa, season_par=Sp)

tt <- seq(0, 3650, by=5)
aa <-seq(0, 365*5, by =5) 
plot(tt, 0.05*F_t(tt), type = "l")
```

To compute anything, we must first set up a function to describe exposure (see the [FoI](FoI.html) vignette). We define functions that plot the FoI for a cohort as it ages (in red), but we can also compute the population average FoI (in black). Different cohorts would experience different histories of exposure. 

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
a3years  = 1:1095
plot(a3years, FoI_a(a3years), type = "l", 
     xlab = "Time or Age (in days)", 
     ylab = expression(z[bday](alpha, a)))
lines(a3years, FoI_a(a3years), col = "darkred", lwd=2)
text(270, 0.02, "Population")
text(100, 0.008, "Cohort", col = "darkred")
```


### Computing `zda` 

The function `ramp.falciparum::zda(alpha, a, FoIpar, ...)` uses the formula in Eq. 1 to compute the density of parasite infections in a cohort of humans as it ages.

Using `zda,` we can compute the density of parasites in a cohort of any age without solving a full system of equations. Given a function describing the FoI in the population, $h(t)$, and the cohort birthday, $d.$

```{r}
# devtools::load_all()
```

```{r zde eg}
alpha = 60
a = 6*365
zda(60, 6*365, FoI_a) 
```

The following computes the density of infections of every age in a cohort of age 3. 

```{r}
zz = zda(a3years, max(a3years), FoI_a)
```

When we plot $z_d(\alpha, a)$, we note that as $\alpha$ grows larger, the parasite cohort gets older. When we plot parasite cohorts by age, time is going backwards on the x-axis. 

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
plot(a3years, zz, type = "l", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(z[bday](alpha,a)))
```


Now, we can imagine what `zda` would look like for several different host cohorts at age three, but who were born at different times. In effect, we are taking a snapshot of the cohorts at the same age, but at different times.

The curves are different because the hosts were born at different months, and they thus experienced different levels of exposure over the first two years of life. Here the annual FoI is 5 infections, per person, per year ($\bar h = 5/365$): 

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
clrs12a <- viridisLite::turbo(12) 
zz = zda(a3years, 1095, FoI_a, bday=90)
plot(a3years, zz, type = "n", ylim = 4 *range(zz), 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(z[bday](alpha,a)))
for(i in 1:12){
  birthday =  30*(i-1) +15
  lines(a3years, zda(a3years, 1095, FoI_a, bday=birthday), col = clrs12a[i]) 
  bp = (0.5 + 2.5*i/12)*365
  points(bp, 0.015, col = clrs12a[i], pch =15)
  text(bp, 0.015, paste(i), col = clrs12a[i], pos=1)
}
  text(650, 0.018, "Birth Month") 
```


## Multiplicity of Infection (MoI) 

We define a random variable $M$ describing the multiplicity of infection (MoI). The distribution of the MoI is Poisson (see the [MoI](MoI.html) vignette).

$$M_d(a) \sim f_M(\zeta; a, d) = \mbox{Pois}(m_d(a))$$

Since $z_d(\alpha, a)$ describes the density of _all_ infections of age $\alpha$ in a cohort of age $a$, the density of all infections must be the MoI. Since $0 \leq \alpha < a$, it must be true that:

\begin{equation}
\tag{2}
m_d(a) = \int_0^a z_d(\alpha, a) d \alpha
\end{equation}

The function that computes $m_d(a)$ is called `meanMoI.` 

```{r}
mm = meanMoI(a3years, FoI_a, hhat=5/365)
```

Here, we plot the average MoI in the host cohort as it ages:


```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
plot(a3years, mm, type = "l",
     xlab = expression(list(a, paste("Host Age (in Days)"))), 
     ylab = expression(m[bday](alpha,a)))
```


## Age of Infection (AoI)

We define a random variable $A_d(a)$ that describes the age of infection (AoI), which is given by the formula 

$$
A_d(a) \sim f_A(\alpha; a, d) = \frac{z_d(\alpha,a)}{m_d(a)}
$$


### The Density Function, `dAoI` 

We can compute $A_d(a)$ using the density function `dAoI`:

```{r}
f_A = dAoI(a3years, max(a3years), FoI_a)
```

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
plot(a3years, f_A, type ="l", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(f[A](alpha, a, bday, h)))
```

Now, as we plot the distribution of the AoI in cohorts at age two, born at different months (as we did above), we notice that the distributions have changed shapes:

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
FA = dAoI(a3years, 1095, FoI_a, bday=70)
plot(a3years, FA, type = "n", ylim = range(FA)*1.15,
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(f[A](alpha, bday)))
for(i in 1:12){
  birthday =  30*(i-1) +15
  lines(a3years, dAoI(a3years, 1095, FoI_a, bday=birthday), col = clrs12a[i]) 
  bp = (0.5 + 2.5*i/12)*365
  points(bp, 0.01, col = clrs12a[i], pch =15)
  text(bp, 0.01, paste(i), col = clrs12a[i], pos=1)
}
  text(650, 0.0115, "Birth Month") 
```


### The Distribution Function, `pAoI` 

The distribution function for $A_d(a)$ is:

$$
F_A(a) \sim \int_0^\alpha f_A(\alpha; a, d) d\alpha
$$
```{r}
F_A = pAoI(a3years, max(a3years), FoI_a)
```

If our functions work correctly, then we should get approximately the same answer from computing the cumulative sum of `dAoI.` 

```{r}
F_A_alt = cumsum(f_A)
```

We shouldn't expect the answers to be exactly the same, but they should be close, with the `pAoI` in black. 

```{r,  fig.height=3.5, fig.width=6}
par(mar = c(5,4,1,1))

plot(a3years, F_A, type = "l", 
     xlab = "Parasite Cohort Age", 
     ylab = expression(1-F[X](alpha, a, bday)), lwd=3)

lines(a3years, F_A_alt, col = "red", lwd=2, lty =2)
```


```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,1))
FA = pAoI(a3years, 1095, FoI_a, bday=70)
plot(a3years, FA, type = "n", ylim = range(FA)*1.15,
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(F[A](alpha)))
for(i in 1:12){
  birthday =  30*(i-1) +15
  lines(a3years, pAoI(a3years, 1095, FoI_a, bday=birthday), col = clrs12a[i]) 
  bp = (0.5 + 2.5*i/12)*365
  points(bp, 0.2, col = clrs12a[i], pch =15)
  text(bp, 0.2, paste(i), col = clrs12a[i], pos=1)
}
text(650, 0.3, "Birth Month") 
```


### Random Numbers, `rAoI`

The function `rAoI` uses `pAoI` to generate random numbers from $F_A(\alpha)$

```{r}
rhx = rAoI(10000, 3*365, FoI_a)
```

A simple visual check computes the empirical CDF for the random variates against $F_A(\alpha)$ computed using `pAoI`

```{r, fig.height=3.5, fig.width=6}
par(mar = c(5,4,1,2))
plot(stats::ecdf(rhx), xlim = c(0,1095), cex=0.2, main = "", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(list(F[A](alpha), paste("ecdf"))))
lines(a3years, F_A, col = "red", lty = 2, lwd=2)
```

We can also plot the distribution functions.

```{r, fig.height=3.5, fig.width=6, echo=F}
par(mar = c(5,4,1,2))
hist(rhx, breaks = seq(0, 1095, by=15), 
     right=F, probability=T, main = "", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     border = grey(0.5))
lines(a3years, f_A, col="red", lwd=2)
```


### AoI Moments 

Let $x$ denote the first moment of of $A_d(a)$: $$x_d(a) = \left< A_d(a) \right> = \int_0^\infty \alpha \frac{z_d(\alpha, a)} {m_d(a)}$$

Similarly, we let $x_d(a)[n]$ denote the higher order moments of $A_d(a)$: 
$$x_{[n]}(a, d) = \int_0^\infty \alpha^n \frac{z_d(\alpha, a)} {m_d(a)}$$

```{r, eval=T}
moment1 = momentAoI(a3years, FoI_a)
moment2 = momentAoI(a3years, FoI_a, n=2)
moment3 = momentAoI(a3years, FoI_a, n=3)
```

The first three moments of the AoY plotted over time. In the top plot, we've also plotted the $n^{th}$ root of the $n^{th}$ moment.

```{r, fig.width=6, fig.height=8, echo=FALSE, eval=T}
par(mfrow = c(3,1), mar = c(5, 4, 0.5, 2))
plot(a3years, moment1, type = "l", xlab = "a - Host Age (in Days)", 
     lwd=2, ylim = range((moment3)^(1/3)),
     ylab = expression(list(x, sqrt(x[paste("[2]")]), sqrt(x[paste("[3]")], 3))))

lines(a3years, sqrt(moment2), col = "darkgreen")
lines(a3years, (moment3)^(1/3), col = "purple")

plot(a3years, moment2, type = "l", xlab = "a - Host Age (in Days)",
     lwd=2, col = "darkgreen",
     ylab = expression(x[paste("[2]")])) 
mar = c(4, 4, 0.5, 2)
plot(a3years, moment3, type = "l", xlab = "a - Host Age (in Days)",
     lwd=2, col = "purple", 
     ylab = expression(x[paste("[3]")])) 
```


## Age of the Youngest Infection (AoY)

We have derived a random variable $Y_d(a)$ describing the age of the youngest infection (AoY). The density function for the AoY is: 

$$Y_d(a) \sim f_Y(\alpha; a, d) = f_A(\alpha, a,d) e^{-m_d(a) F_A(\alpha, a,d)} \frac{m_d(a)}{p_d(a)}$$ 
The distribution function is: 

$$F_Y(a) \sim
\frac{1-e^{-m_d(a)F_A(\alpha, a,d)}}{1-e^{-m_d(a)}} = 
\frac{1-e^{-m_d(a)F_A(\alpha, a,d)}}{p_d(a)} 
$$

The derivations are found in a Suppplement to Henry JM, *et al.* (2024).

The mean AoY is: 

$$
\left< Y_d(a) \right> = \int_0^a \alpha \; f_Y(\alpha, a, d) \; d\alpha $$ 

And the higher order moments for the AoY are: 

$$\left< Y_d(a)^n \right> = \int_0^n \alpha^n \; f_y(\alpha, a, d) \; d\alpha $$ 



### AoY Density, `dAoY`  

The density function is computed with the function `dAoY.`

```{r, eval=T}
f_Y = dAoY(a3years, 3*365, FoI_a)
```

We can compare $f_Y(\alpha)$ (in black) to $f_A(\alpha)$ (in grey). 

```{r, fig.height=3.5, fig.width=6, echo=F, eval=T}
par(mar = c(5,4,1,1))
plot(a3years, f_Y, type ="l", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     ylab = expression(list(f[Y](alpha), f[A](alpha))), ylim = range(f_Y), lwd=2)
lines(a3years, f_A, col = grey(0.5))
```


### Random Variables, `rAoY`  

```{r, eval=T}
raoy = rAoY(10^5, 3*365, FoI_a)
```

```{r, fig.height=4, fig.width=7, eval=T} 
hist(raoy, breaks=seq(0, 1095, by = 15), 
     right=F, probability=T, main = "", 
     xlab = expression(list(alpha, paste("Parasite Age (in Days)"))), 
     border = grey(0.5)) -> out
lines(a3years, f_Y, type = "l", col = "red") 
```

### AoY Moments

```{r, eval=T}
aa = seq(5, 3*365, by = 5) 
moment1y = momentAoY(aa, FoI_a)
moment2y = momentAoY(aa, FoI_a, n=2)
moment3y = momentAoY(aa, FoI_a, n=3)
```

The first three moments of the AoY plotted over time. In the top plot, we've also plotted the $n^{th}$ root of the $n^{th}$ moment.

```{r, fig.width=6, fig.height=8, echo=FALSE, eval=T}
par(mfrow = c(4,1), mar = c(0.5, 4, 0.5, 2))
plot(aa, moment1y, type = "l", xlab = "", ylab = expression(E(Y)), lwd=2, xaxt="n", ylim = range((moment3y)^(1/3)) ) 
lines(aa, sqrt(moment2y), col = "darkgreen")
lines(aa, (moment3y)^(1/3), col = "purple")
plot(aa, moment2y, type = "l", xlab = "", ylab = expression(E(Y^2)), lwd=2, xaxt="n", col = "darkgreen") 
plot(aa, moment3y, type = "l", xlab = "", ylab = expression(E(Y^2)), lwd=2, col = "purple") 
mtext("Age (in Days)", 1, 3)
```

*** 

**Next:** 

+ In the vignette [MoI](MoI.html), we show that the mean MoI computed using the formula in Eq. 2 gives the same answer as other approaches. 

