---
title: "The SIP Hybrid Model"
subtitle: "Comparing Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The SIP Hybrid Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This describes a queuing model `SIP-moi`

> see `SIP.R` 


```{r}
devtools::load_all()
```

The system we are interested in modeling describes chemoprotected ($P$) and susceptible ($S$) populations. 

\begin{equation}
    \begin{array}{rl}
    \frac{dP}{da} &= - \eta P + (h \rho + \xi) \left( S + I \right) + \sigma I  -\mu P\\ 
    \frac{dS}{da} &= \eta P - r I_1  - (h + \mu + \xi) S
    \end{array}
\end{equation}

We track the number infected with $i$ types, $I_i.$

\begin{equation}
    \begin{array}{rl}
    \frac{dI_1}{da} &=  2r I_2  + h (1-\rho) S - (r + h + \sigma + \xi + \mu ) I_1 \\ 
    \frac{dI_2}{da} &=  3r I_3 + h (1-\rho) I_1 - (2r + h + \sigma + \xi + \mu) I_2 \\ 
    \end{array}
\end{equation}

The implementation truncates the infinite system, choosing a largest MoI, $n$: 

\begin{equation}
\frac{dI_n}{da} = h (1-\rho) I_{n-1} - (n r + h \rho + \sigma + \xi+ \mu) I_n
\end{equation}


We let $$m = \sum_i \frac{I_i}H$$ denote the mean MoI in the population. We can show that
$$ 
\frac{\textstyle{dm}}{\textstyle{da}} = h(1-\rho) s - \left(r + \rho h + \sigma + \xi  \right)  m
$$
The function `solveSIPmoi`:

+ calls `d_SIPmoi_da` which computes all the derivaves for $S$, $I$, $P$, and $m$

+ computes $m$ from $I$ 

```{r, fig.width=7}
foiP1 = list(hbar = 1, agePar = par_flatAge(), seasonPar = par_flatSeason(), trendPar = par_flatTrend())
out <- solveSIPmoi(2/365, mu=0, rho=0, foiP1, Amax=5*365)
```

We note that the variable $m$ matches the MoI computed from $I$: 

```{r, fig.width=7}
with(out, plot(age, moi, type ="l"))
with(out, lines(age, m, col = "red", lty=2))
```

The change in prevalence with respect to age is: 

```{r, fig.width=7}
with(out, plot(age, x, type ="l"))
```
```{r, fig.height=4, fig.width=7}
with(out, plot(age, 0.005*dnbinom(1,mu=m, size=cv)/x, col = "red", type = "l", ylim = c(0, 0.0055), ylab = expression(F[r](m))))
with(out, lines(age, Fr, type ="l"))
```



```{r, fig.width=7}
with(out, plot(age, cv, type ="l"))
```


```{r, fig.width=7}
with(out, plot(age, var, type ="l"))
```


```{r, fig.width=7}
with(out, plot(1:N+0, II[730,], type ="h", lwd=2, xlab="MoI", ylim = c(0,1000)))
with(out, lines(1:N+0.1, II[1095,], type ="h", lwd=2))
with(out, lines(1:N+0.2, II[1460,], type ="h", lwd=2))
```


